---
output:
  md_document:
    variant: markdown_github
references:
- id: Touloumis2016
  title: "Simulating Correlated Binary and Multinomial Responses under Marginal Model Specification: The SimCorMultRes Package"
  author:
  - family: Touloumis
    given: Anestis
  container-title: The R Journal
  volume: 8
  URL: 'https://journal.r-project.org/archive/2016/RJ-2016-034/index.html'
  issue: 2
  page: 79-91
  type: article-journal
  issued:
    year: 2016
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
library("badger")
knitr::opts_chunk$set(
  tidy= TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# SimCorMultRes: Simulates Correlated Multinomial Responses


[![Travis-CI Build Status](https://travis-ci.org/AnestisTouloumis/SimCorMultRes.svg?branch=master)](https://travis-ci.org/AnestisTouloumis/SimCorMultRes)
`r badge_devel("AnestisTouloumis/SimCorMultRes", "brightgreen")`
[![Project Status: Active The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active) 
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](/commits/master)

[![CRAN Version](http://www.r-pkg.org/badges/version/SimCorMultRes?color=blue)](https://cran.r-project.org/package=SimCorMultRes)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/grand-total/SimCorMultRes?color=blue)](http://cranlogs.r-pkg.org/badges/grand-total/SimCorMultRes)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/SimCorMultRes)](http://cran.rstudio.com/web/packages/SimCorMultRes/index.html)


## Author

Anestis Touloumis <https://brighton.academia.edu/AnestisTouloumis>

School of Computing, Engineering and Mathematics, University of Brighton. 


## Installation

You can install the release version of the **SimCorMultRes** package:

```{r eval=FALSE}
install.packages("SimCorMultRes")
```

The source code for the release version of **SimCorMultRes** is available on CRAN at:

- https://CRAN.R-project.org/package=SimCorMultRes

Or you can install the development version of the **SimCorMultRes** package:

```{r eval=FALSE}
install.packages("devtools") # if you have not installed "devtools" package
devtools::install_github("AnestisTouloumis/SimCorMultRes")
```

The source code for the development version of **SimCorMultRes** is available on github at:

- https://github.com/AnestisTouloumis/SimCorMultRes.

To use **SimCorMultRes**, you should load the package as follows:

```{r}
library(SimCorMultRes)
```

## Usage and functions

The **SimCorMultRes** package provides functions to simulate correlated binary or multinomial responses. These responses are drawn as realizations of a latent regression model for continuous random vectors with the correlation structure expressed in terms of the latent correlation [@Touloumis2016].


The package contains four core functions:

- `rbin` to simulate correlated binary responses,
- `rmult.bcl` to simulate correlated nominal multinomial responses,
- `rmult.clm` to simulate correlated ordinal responses under a marginal cumulative link model,
- `rmult.clm` to simulate correlated ordinal responses under a marginal continuation-ratio link model.


The **SimCorMultRes** package also offers two utility functions:

- `rnorta` for simulating random vectors using the NORTA method,
- `rsmvnorm` for simulating continuous random vectors from a multivariate normal distribution.


## Example

The following R code illustrates how to use the core function `rbin`:

```{r}
## See Example 3.4 in the Vignette.
set.seed(123)
N <- 5000
clsize <- 4
intercepts <- 0
betas <- 0.2
cor.matrix <- toeplitz(c(1, 0.9, 0.9, 0.9))
x <- rep(rnorm(N), each = clsize)
CorBinRes <- rbin(clsize = clsize, intercepts = intercepts, betas = betas, 
    xformula = ~x, cor.matrix = cor.matrix, link = "probit")
library(gee)
binGEEmod <- gee(y ~ x, family = binomial("probit"), id = id, data = CorBinRes$simdata)
summary(binGEEmod)$coefficients
```

Additional examples can be found in @Touloumis2016 and in the vignette of **SimCorMultRes**.

## Getting help
A detailed description of **SimCorMultRes** can be found in @Touloumis2016. The vignette of **SimCorMultRes** contains additional examples and it can be accessed by executing the following R commands: 

```{r, eval=FALSE}
browseVignettes("SimCorMultRes")
```


## How to cite

```{r}
citation("SimCorMultRes")
```

# References