---
output: github_document
references:
- id: Touloumis2016
  title: "Simulating Correlated Binary and Multinomial Responses under Marginal Model Specification: The SimCorMultRes Package"
  author:
  - family: Touloumis
    given: Anestis
  container-title: The R Journal
  volume: 8
  URL: 'https://journal.r-project.org/archive/2016/RJ-2016-034/index.html'
  issue: 2
  page: 79-91
  type: article-journal
  issued:
    year: 2016
csl: biometrics.csl    
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 80),
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# SimCorMultRes: Simulates Correlated Multinomial Responses

[![Github version](`r paste0("https://img.shields.io/badge/GitHub%20-", as.vector(read.dcf('DESCRIPTION')[, 'Version']),"-orange.svg")`)]("commits/master")
[![R-CMD-check](https://github.com/AnestisTouloumis/SimCorMultRes/workflows/R-CMD-check/badge.svg)](https://github.com/AnestisTouloumis/SimCorMultRes/actions)
[![Project Status: Active The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active) 
[![Codecov test coverage](https://codecov.io/gh/AnestisTouloumis/SimCorMultRes/branch/master/graph/badge.svg)](https://codecov.io/gh/AnestisTouloumis/SimCorMultRes?branch=master)

[![CRAN Version](https://www.r-pkg.org/badges/version/SimCorMultRes?color=blue)](https://cran.r-project.org/package=SimCorMultRes)
[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/grand-total/SimCorMultRes?color=blue)](https://cranlogs.r-pkg.org/badges/grand-total/SimCorMultRes)
[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/SimCorMultRes)](https://cran.r-project.org/package=SimCorMultRes)


## Installation

You can install the release version of `SimCorMultRes`:

```{r eval=FALSE}
install.packages("SimCorMultRes")
```

The source code for the release version of `SimCorMultRes` is available on CRAN at:

- https://CRAN.R-project.org/package=SimCorMultRes

Or you can install the development version of `SimCorMultRes`:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("AnestisTouloumis/SimCorMultRes")
```

The source code for the development version of `SimCorMultRes` is available on github at:

- https://github.com/AnestisTouloumis/SimCorMultRes

To use `SimCorMultRes`, you should load the package as follows:

```{r}
library("SimCorMultRes")
```

## Usage and functions

This package provides five core functions to simulate correlated binary (`rbin`), nominal (`rmult.bcl`) and ordinal (`rmult.acl`, `rmult.clm` and `rmult.crm`) responses, which are drawn as realizations of a latent regression model for continuous random vectors as proposed by @Touloumis2016:

- `rbin` to simulate correlated binary responses under a marginal model with logit, probit, cloglog and cauchit link function,
- `rmult.bcl` to simulate correlated nominal multinomial responses under a marginal baseline-category logit model,
- `rmult.acl` to simulate correlated ordinal responses under a marginal adjacent-category logit model,
- `rmult.clm` to simulate correlated ordinal responses under a marginal cumulative link model,
- `rmult.crm` to simulate correlated ordinal responses under a marginal continuation-ratio link model.

All five functions, assume that you provide either the correlation matrix of the multivariate normal distribution in NORTA (via `cor.matrix`) or the latent responses (via the `rlatent`). 

There are also two utility functions:

- `rnorta` for simulating continuous or discrete random vectors with prescribed marginal distributions using the NORTA method,
- `rsmvnorm` for simulating continuous random vectors from a multivariate normal distribution.


## Example

The following R code illustrates how to use the core function `rbin`:

```{r}
## See Example 3.5 in the Vignette.
set.seed(123)
## define number of random vectors
sample_size <- 100
## define size of each random vector
cluster_size <- 4
## define intercept of the binary probit regression model
beta_intercepts <- 0
## define coefficients of the explanatory variables
beta_coefficients <- 0.2
## provide explanatory variables
x <- rep(rnorm(sample_size), each = cluster_size)
## define correlation matrix for the multivariate normal distribution in NORTA
latent_correlation_matrix <- toeplitz(c(1, 0.9, 0.9, 0.9))
## use rbin function to create the desired dataset
simulated_binary_responses <- rbin(clsize = cluster_size,
                                   intercepts = beta_intercepts,
                                   betas = beta_coefficients, xformula = ~  x,
                                   cor.matrix = latent_correlation_matrix,
                                   link = "probit")
library("gee")
binary_gee_model <- gee(y ~ x, family = binomial("probit"), id = id,
                        data = simulated_binary_responses$simdata)
summary(binary_gee_model)$coefficients
```

Additional examples can be found in @Touloumis2016 and in the vignette of `SimCorMultRes`. To access these two documents, run the following command:

```{r eval=FALSE}
browseVignettes("SimCorMultRes")
```

## How to cite

```{r echo=FALSE, comment=""}
citation("SimCorMultRes")
```

# References
