---
title: "Simulating Correlated Binary and Multinomial Responses with SimCorMultRes"
author: "Anestis Touloumis"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  bookdown::html_document2:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Simulating Correlated Binary and Multinomial Responses}
  %\VignetteKeywords{correlated binary responses, correlated multinomial responses, correlated ordinal responses, correlated nominal responses, GEE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "SimCorMultResBib.bib"
biblio-style: "apalike"
link-citations: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  collapse = TRUE, 
  comment = "#>"
  )
```

# Introduction 
The R package `SimCorMultRes` is suitable for simulation of correlated binary responses (exactly two response categories) and of correlated nominal or ordinal multinomial responses (three or more response categories) conditional on a regression model specification for the marginal probabilities of the response categories. This vignette briefly describes the simulation methods proposed by @Touloumis2016 and illustrates the use of the core functions of `SimCorMultRes`. A more detailed description of `SimCorMultRes` can be found in @Touloumis2016. 

# Areas of Applications
This package was created to facilitate the task of carrying out simulation studies and evaluating the performance of statistical methods for estimating the regression parameters in a marginal model with clustered binary and multinomial responses. Examples of such statistical methods include maximum likelihood methods, copula approaches, quasi-least squares approaches, generalized quasi-likelihood methods and generalized estimating equations (GEE) approaches among others [see references in @Touloumis2016]. 

In addition, `SimCorMultRes` can generate correlated binary and multinomial random variables conditional on a desired dependence structure and known marginal probabilities even if these are not determined by a regression model [see third example in @Touloumis2016] or to explore approximations of association measures for discrete variables that arise as realizations of an underlying continuum [see second example in @Touloumis2016].

# Simulation Methods
Let $Y_{it}$ be the binary or multinomial response for subject $i$ ($i=1,\ldots,N$) at measurement occasion $t$ ($t=1,\ldots,T$), and let $\mathbf {x}_{it}$ be the associated covariates vector. We assume that $Y_{it} \in \{0,1\}$ for binary responses and $Y_{it} \in \{1,2,\ldots,J\geq 3\}$ for multinomial responses. 

## Correlated nominal responses 
The function `rmult.bcl` simulates nominal responses under the marginal baseline-category logit model
\begin{equation}
\log \left[\frac{\Pr(Y_{it}=j |\mathbf {x}_{it})}{\Pr(Y_{it}=J |\mathbf {x}_{it})}\right]=(\beta_{tj0}-\beta_{tJ0})+(\boldsymbol {\beta}_{tj}-\boldsymbol{\beta}_{tJ})^{\prime} \mathbf {x}_{it}=\beta^{\ast}_{tj0}+\boldsymbol{\beta}^{\ast\prime}_{tj}\mathbf {x}_{it},
(\#eq:MBCLM)
\end{equation}
where $\beta_{tj0}$ is the $j$-th category-specific intercept at measurement occasion $t$ and $\boldsymbol{\beta}_{tj}$ is the $j$-th category-specific parameter vector associated with the covariates at measurement occasion $t$. The popular identifiability constraints $\beta_{tJ0}=0$ and $\boldsymbol{\beta}_{tJ}=\mathbf {0}$ for all $t$, imply that $\beta^{\ast}_{tj0}=\beta_{tj0}$ and $\boldsymbol {\beta}^{\ast}_{tj}=\boldsymbol{\beta}_{tj}$ for all $t=1,\ldots,T$ and $j=1,\ldots,J-1$. The threshold 
$$Y_{it}=j \Leftrightarrow U^{NO}_{itj}=\max \{U^{NO}_{it1},\ldots,U^{NO}_{itJ}\}$$
generates clustered nominal responses that satisfy the marginal baseline-category logit model \@ref(eq:MBCLM), where
$$U^{NO}_{itj}=\beta_{tj0}+\boldsymbol{\beta}_{tj}^{\prime} \mathbf {x}_{it}+e^{NO}_{itj},$$
and where the random variables $\{e^{NO}_{itj}:i=1,\ldots,N \text{,  } t=1,\ldots,T \text{ and } j=1,\ldots,J\}$ satisfy the following conditions: 

1. $e^{NO}_{itj}$ follows the standard extreme value distribution for all $i$, $t$ and $j$ (mean $=\gamma \approx 0.5772$, where $\gamma$ is Euler's constant, and variance $=\pi^2/6$).
1. $e^{NO}_{i_1t_1j_1}$ and $e^{NO}_{i_2t_2j_2}$ are independent random variables provided that $i_1 \neq i_2$.
1. $e^{NO}_{itj_1}$ and $e^{NO}_{itj_2}$ are independent random variables provided that $j_1\neq j_2$.

For each subject $i$, the association structure among the clustered nominal responses $\{Y_{it}:t=1,\ldots,T\}$ depends on the joint distribution and correlation matrix of $\{e^{NO}_{itj}:t=1,\ldots,T \text{ and } j=1,\ldots,J\}$. If the random variables $\{e^{NO}_{itj}:t=1,\ldots,T \text{ and } j=1,\ldots,J\}$ are independent then so are $\{Y_{it}:t=1,\ldots,T\}$.
\newline{}

```{example, Example1, name="Simulation of clustered nominal responses using the NORTA method"}
Suppose the aim is to simulate nominal responses from the marginal baseline-category logit model
\begin{equation*}
\log \left[\frac{\Pr(Y_{it}=j |\mathbf {x}_{it})}{\Pr(Y_{it}=4 |\mathbf {x}_{it})}\right]=\beta_{j0}+ \beta_{j1} {x}_{i1}+ \beta_{j2} {x}_{it2} \end{equation*}
where $N=500$, $T=3$, $(\beta_{10},\beta_{11},\beta_{12},\beta_{20},\beta_{21},\beta_{22},\beta_{30},\beta_{31},\beta_{32})=(1, 3, 2, 1.25, 3.25, 1.75, 0.75, 2.75, 2.25)$ and $\mathbf {x}_{it}=(x_{i1},x_{it2})^{\prime}$ for all $i$ and $t$, with $x_{i1}\overset{iid}{\sim} N(0,1)$ and $x_{it2}\overset{iid}{\sim} N(0,1)$. For the dependence structure, suppose that the correlation matrix $\mathbf{R}$ in the NORTA method has elements 
\[
\mathbf{R}_{t_1j_1,t_2j_2}=\begin{cases}
1 & \text{if } t_1=t_2 \text{ and } j_1=j_2\\
0.95 & \text{if } t_1 \neq t_2 \text{ and } j_1=j_2\\
0 & \text{otherwise }\\
\end{cases}
\]
for all $i=1,\ldots,500$.
```

```{r, tidy=TRUE}
# parameter vector
betas <-  c(1, 3, 2, 1.25, 3.25, 1.75, 0.75, 2.75, 2.25, 0, 0, 0)
# sample size
N <- 500 
# number of nominal response categories
ncategories <- 4  
# cluster size
clsize <- 3 
set.seed(1) 
# time-stationary covariate x_{i1}
x1 <- rep(rnorm(N), each = clsize)
# time-varying covariate x_{it2}
x2 <- rnorm(N * clsize) 
# create covariates dataframe 
xdata <- data.frame(x1, x2) 
set.seed(321)
library(SimCorMultRes)
# correlation matrix for the NORTA method
cor.matrix <- kronecker(toeplitz(c(1,rep(0.95,clsize-1))),diag(ncategories)) 
# simulation of clustered nominal responses
CorNomRes <- rmult.bcl(clsize = clsize, ncategories = ncategories, betas = betas, xformula = ~x1 + x2, xdata = xdata, cor.matrix = cor.matrix)
suppressPackageStartupMessages(library("multgee"))
# fitting a GEE model
fit <- nomLORgee(y~x1+x2,data=CorNomRes$simdata,id=id,repeated=time,LORstr="time.exch")
# checking regression coefficients
round(coef(fit),2)
```
## Correlated ordinal responses
Simulation of clustered ordinal responses is feasible under either a marginal cumulative link model or a marginal continuation-ratio model.

### Marginal cumulative link model
The function `rmult.clm` simulates ordinal responses under the marginal cumulative link model
\begin{equation}
\Pr(Y_{it}\le j |\mathbf {x}_{it})=F(\beta_{tj0} +\boldsymbol {\beta}^{\prime}_t \mathbf {x}_{it})
(\#eq:MCLM)
\end{equation}
where $F$ is a cumulative distribution function (cdf), $\beta_{tj0}$ is the $j$-th category-specific intercept at measurement occasion $t$ and $\boldsymbol \beta_t$ is the regression parameter vector associated with the covariates at measurement occasion $t$. The category-specific intercepts at each measurement occasion $t$ are assumed to be monotone increasing, that is
$$-\infty=\beta_{t00} <\beta_{t10} < \beta_{t20} < \cdots < \beta_{t(J-1)0}< \beta_{tJ0}=\infty$$
for all $t$. Using the threshold
$$Y_{it}=j \Leftrightarrow \beta_{t(j-1)0} < U^{O1}_{it} \leq \beta_{tj0}$$
clustered ordinal responses that satisfy the marginal cumulative link model \@ref(eq:MCLM) are generated, where
$$U^{O1}_{it}=-\boldsymbol{\beta}^{\prime}_t \mathbf {x}_{it}+e^{O1}_{it},$$
and where $\{e^{O1}_{it}:i=1,\ldots,N \text{ and } t=1,\ldots,T\}$ are random variables such that:
 
1. $e^{O1}_{it} \sim F$ for all $i$ and $t$.
1. $e^{O1}_{i_1t_1}$ and $e^{O1}_{i_2t_2}$ are independent random variables provided that $i_1 \neq i_2$. 

For each subject $i$, the association structure among the clustered ordinal responses $\{Y_{it}:t=1,\ldots,T\}$ depends on the pairwise bivariate distributions and correlation matrix of $\{e^{O1}_{it}:t=1,\ldots,T\}$. If the random variables $\{e^{O1}_{it}:t=1,\ldots,T\}$ are independent then so are $\{Y_{it}:t=1,\ldots,T\}$.
\newline{} 
```{example, name="Simulation of clustered ordinal responses conditional on a marginal cumulative probit model with time-varying regression parameters"}
Suppose the goal is to simulate correlated ordinal responses from the marginal cumulative probit model
\begin{equation*}
\Pr(Y_{it}\le j |\mathbf x_{it})=\Phi(\beta_{j0} + \beta_{t1} {x}_{i})
\end{equation*}
where $\Phi$ denotes the cdf of the standard normal distribution (mean $=0$ and variance $=1$), $N=500$, $T=4$, $(\beta_{10},\beta_{20},\beta_{30},\beta_{40})=(-1.5,-0.5,0.5,1.5)$, $(\beta_{11},\beta_{21},\beta_{31},\beta_{41})=(1,2,3,4)$ and $\mathbf x_{it}=x_{i}\overset{iid}{\sim} N(0,1)$ for all $i$ and $t$. For the dependence structure, assume that $\mathbf{e}_i^{O1}=(e^{O1}_{i1},e^{O1}_{i2},e^{O1}_{i3},e^{O1}_{i4})^{\prime}$ are iid random vectors from a tetra-variate normal distribution with mean vector the zero vector and covariance matrix the correlation matrix
$$\left( {\begin{array}{*{20}c}
   1.00 & 0.85 & 0.50 & 0.15  \\
   0.85 & 1.00 & 0.85 & 0.50  \\
   0.50 & 0.15 & 1.00 & 0.85 \\
   0.15 & 0.85 & 0.50 & 1.00
 \end{array} } \right).$$
```

```{r}
set.seed(12345)
# sample size
N <- 500
# cluster size
clsize <- 4
# category-specific intercepts
intercepts <- c(-1.5, -0.5, 0.5, 1.5)
# time-varying regression parameters associated with covariates
betas <- matrix(c(1, 2, 3, 4), 4, 1)
# time-stationary covariate 
x <- rep(rnorm(N), each = clsize)
# correlation matrix for the NORTA method
cor.matrix <- toeplitz(c(1, 0.85, 0.5, 0.15))
# simulation of ordinal responses
CorOrdRes <- rmult.clm(clsize = clsize, intercepts = intercepts, betas = betas, 
    xformula = ~x, cor.matrix = cor.matrix, link = "probit")
# first eight rows of the simulated dataframe
head(CorOrdRes$simdata, n = 8)
```

### Marginal continuation-ratio model
The function `rmult.crm` simulates clustered ordinal responses under the marginal continuation-ratio model
\begin{equation}
\Pr(Y_{it}=j |Y_{it} \ge j,\mathbf {x}_{it})=F(\beta_{tj0} +\boldsymbol {\beta}^{'}_t \mathbf {x}_{it})
(\#eq:MCRM)
\end{equation}
where $\beta_{tj0}$ is the $j$-th category-specific intercept at measurement occasion $t$, $\boldsymbol \beta_t$ is the regression parameter vector associated with the covariates at measurement occasion $t$ and $F$ is a cdf. This is accomplished by utilizing the threshold
$$Y_{it}=j, \text{ given } Y_{it} \geq j \Leftrightarrow U^{O2}_{itj} \leq \beta_{tj0}$$
where
$$U^{O2}_{itj}=-\boldsymbol {\beta}^{\prime}_t \mathbf {x}_{it}+e^{O2}_{itj},$$
and where $\{e^{O2}_{itj}:i=1,\ldots,N \text{ , } t=1,\ldots,T \text{ and } j=1,\ldots,J-1\}$ satisfy the following three conditions:
 
1. $e^{O2}_{itj} \sim F$ for all $i$, $t$ and $j$.
1. $e^{O2}_{i_1t_1j_1}$ and $e^{O2}_{i_2t_2j_2}$ are independent random variables provided that $i_1 \neq i_2$.
1. $e^{O2}_{itj_1}$ and $e^{O2}_{itj_2}$ are independent random variables provided that $j_1\neq j_2$.
 
For each subject $i$, the association structure among the clustered ordinal responses $\{Y_{it}:t=1,\ldots,T\}$ depends on the joint distribution and correlation matrix of $\{e^{O2}_{itj}:j=1,\ldots,J \text{ and } t=1,\ldots,T\}$. If the random variables $\{e^{O2}_{itj}:j=1,\ldots,J \text{ and } t=1,\ldots,T\}$ are independent then so are $\{Y_{it}:t=1,\ldots,T\}$. 
\newline{}
```{example,name="Simulation of clustered ordinal responses conditional on a marginal continuation-ratio probit model"}
Suppose simulation of clustered ordinal responses under the marginal continuation-ratio probit model
\begin{equation*}
\Pr(Y_{it}=j |Y_{it} \ge j,\mathbf{x}_{it})=\Phi(\beta_{j0} + \beta {x}_{it})
\end{equation*}
with $N=500$, $T=4$, $(\beta_{10},\beta_{20},\beta_{30},\beta_{40},\beta)=(-1.5,-0.5,0.5,1.5,1)$ and $\mathbf{x}_{it}=x_{it}\overset{iid}{\sim} N(0,1)$ for all $i$ and $t$ is desired. For the dependence structure, assume that $\left\{\mathbf{e}_i^{O2}=\left(e^{O2}_{i11},\ldots,e^{O1}_{i44}\right)^{\prime}:i=1,\ldots,N\right\}$ are iid random vectors from a multivariate normal distribution with mean vector the zero vector and covariance matrix the $16 \times 16$ correlation matrix with elements\[
    \text{corr}(e^{O2}_{it_1j_1},e^{O2}_{it_2j_2}) = \begin{cases}
        1  & \text{for } j_1 = j_2 \text{ and } t_1 = t_2\\
        0.24  & \text{for } t_1 \neq t_2\\
        0  & \text{otherwise.}\\
        \end{cases} 
  \]
```

```{r}
set.seed(1)
# sample size
N <- 500
# cluster size
clsize <- 4
# category-specific intercepts
intercepts <- c(-1.5, -0.5, 0.5, 1.5)
# regression parameters associated with covariates
betas <- 1
# time-varying covariate
x <- rnorm(N * clsize)
# number of ordinal response categories
ncategories <- 5
# correlation matrix for the NORTA method
cor.matrix <- diag(1, (ncategories - 1) * clsize) + kronecker(toeplitz(c(0, 
    rep(0.24, ncategories - 2))), matrix(1, clsize, clsize))
# simulation of ordinal responses
CorOrdRes <- rmult.crm(clsize = clsize, intercepts = intercepts, betas = betas, 
    xformula = ~x, cor.matrix = cor.matrix, link = "probit")
# first six clusters with ordinal responses
head(CorOrdRes$Ysim)
```

### Marginal adjacent-category logit model
The function `rmult.acl` simulates clustered ordinal responses under the marginal adjacent-category logit model
\begin{equation}
\log\left[\frac{\Pr(Y_{it}=j |\mathbf {x}_{it})}{\Pr(Y_{it}=j+1 |\mathbf {x}_{it})}\right]=\beta_{tj0} +\boldsymbol {\beta}^{'}_t \mathbf {x}_{it}
(\#eq:MACL)
\end{equation}
where $\beta_{tj0}$ is the $j$-th category-specific intercept at measurement occasion $t$, $\boldsymbol \beta_t$ is the regression parameter vector associated with the covariates at measurement occasion $t$. 

Generation of clustered ordinal responses relies upon utilizing the connection between baseline-category logit models and adjacent-category logit models. In particular, the threshold 
$$Y_{it}=j \Leftrightarrow U^{O3}_{itj}=\max \{U^{O3}_{it1},\ldots,U^{O3}_{itJ}\}$$
generates clustered nominal responses that satisfy the marginal adjacent-category logit model \@ref(eq:MACL), where
$$U^{O3}_{itj}=\sum_{k=j}^J\beta_{tk0}+(J-j)\boldsymbol{\beta}_{t}^{\prime} \mathbf {x}_{it}+e^{O3}_{itj},$$
and where the random variables $\{e^{O3}_{itj}:i=1,\ldots,N \text{,  } t=1,\ldots,T \text{ and } j=1,\ldots,J\}$ satisfy the following conditions: 

1. $e^{O3}_{itj}$ follows the standard extreme value distribution for all $i$, $t$ and $j$.
1. $e^{O3}_{i_1t_1j_1}$ and $e^{O3}_{i_2t_2j_2}$ are independent random variables provided that $i_1 \neq i_2$.
1. $e^{O3}_{itj_1}$ and $e^{O3}_{itj_2}$ are independent random variables provided that $j_1\neq j_2$.

For each subject $i$, the association structure among the clustered ordinal responses $\{Y_{it}:t=1,\ldots,T\}$ depends on the joint distribution and correlation matrix of $\{e^{O3}_{itj}:t=1,\ldots,T \text{ and } j=1,\ldots,J\}$. If the random variables $\{e^{O3}_{itj}:t=1,\ldots,T \text{ and } j=1,\ldots,J\}$ are independent then so are $\{Y_{it}:t=1,\ldots,T\}$.
\newline{} 

```{example, Example4, name="Simulation of clustered ordinal responses conditional on a marginal adjacent-category logit model using the NORTA method"}
Suppose the aim is to simulate ordinal responses from the marginal adjacent-category logit model
\begin{equation*}
\log \left[\frac{\Pr(Y_{it}=j |\mathbf {x}_{it})}{\Pr(Y_{it}=j+1 |\mathbf {x}_{it})}\right]=\beta_{j0}+ \beta_{1} {x}_{i1}+ \beta_{2} {x}_{it2} \end{equation*}
where $N=500$, $T=3$, $(\beta_{10},\beta_{20},\beta_{30})=(3, 2, 1)$, $(\beta_{1},\beta_{2})=(1, 1)$ and $\mathbf {x}_{it}=(x_{i1},x_{it2})^{\prime}$ for all $i$ and $t$, with $x_{i1}\overset{iid}{\sim} N(0,1)$ and $x_{it2}\overset{iid}{\sim} N(0,1)$. For the dependence structure, suppose that the correlation matrix $\mathbf{R}$ in the NORTA method has elements 
\[
\mathbf{R}_{t_1j_1,t_2j_2}=\begin{cases}
1 & \text{if } t_1=t_2 \text{ and } j_1=j_2\\
0.95 & \text{if } t_1 \neq t_2 \text{ and } j_1=j_2\\
0 & \text{otherwise }\\
\end{cases}
\]
for all $i=1,\ldots,500$.
```

```{r, tidy=TRUE}
# intercepts
intercepts <- c(3, 2, 1)
# parameter vector
betas <- c(1, 1)
# sample size
N <- 500 
# cluster size
clsize <- 3 
set.seed(321) 
# time-stationary covariate x_{i1}
x1 <- rep(rnorm(N), each = clsize)
# time-varying covariate x_{it2}
x2 <- rnorm(N * clsize) 
# create covariates dataframe 
xdata <- data.frame(x1, x2) 
# correlation matrix for the NORTA method
cor.matrix <- kronecker(toeplitz(c(1,rep(0.95,clsize-1))),diag(4)) 
# simulation of clustered ordinal responses
CorNorRes <- rmult.acl(clsize = clsize, intercepts = intercepts, betas = betas, xformula = ~x1 + x2, xdata = xdata, cor.matrix = cor.matrix)
suppressPackageStartupMessages(library("multgee"))
# fitting a GEE model
fit <- ordLORgee(y~x1+x2,data=CorNorRes$simdata,id=id,repeated=time,LORstr="time.exch", link = "acl")
# checking regression coefficients
round(coef(fit),2)
```

## Correlated binary responses
The function `rbin` simulates binary responses under the marginal model specification
\begin{equation}
\Pr(Y_{it}=1 |\mathbf {x}_{it})=F(\beta_{t0} +\boldsymbol {\beta}^{\prime}_{t} \mathbf {x}_{it})
(\#eq:MB)
\end{equation}
where $\beta_{t0}$ is the intercept at measurement occasion $t$, $\boldsymbol \beta_t$ is the regression parameter vector associated with the covariates at measurement occasion $t$ and $F$ is a cdf. The threshold
$$Y_{it}=1 \Leftrightarrow U^{B}_{it} \leq \beta_{t0} + 2 \boldsymbol {\beta}^{\prime}_t \mathbf {x}_{it},$$
generates clustered binary responses that satisfy the marginal model \@ref(eq:MB), where
\begin{equation}
U^{B}_{it}=\boldsymbol {\beta}^{\prime}_t \mathbf {x}_{it}+e^{B}_{it},
(\#eq:TMB)
\end{equation}
and where $\{e^{B}_{it}:i=1,\ldots,N \text{ and } t=1,\ldots,T\}$ are random variables such that:

1. $e^{B}_{it} \sim F$ for all $i$ and $t$.
1. $e^{B}_{i_1t_1}$ and $e^{B}_{i_2t_2}$ are independent random variables provided that $i_1 \neq i_2$.

For each subject $i$, the association structure among the clustered binary responses $\{Y_{it}:t=1,\ldots,T\}$ depends on the pairwise bivariate distributions and correlation matrix of $\{e^{B}_{it}:t=1,\ldots,T\}$. If the random variables $\{e^{B}_{it}:t=1,\ldots,T\}$ are independent then so are $\{Y_{it}:t=1,\ldots,T\}$.
\newline{}
```{example, label=BinaryProbit, name="Simulation of clustered binary responses conditional on a marginal probit model using NORTA method"}
Suppose the goal is to simulate clustered binary responses from the marginal probit model 
$$\Pr(Y_{it}=1 |\mathbf{x}_{it})=\Phi(0.2x_i)$$
where $N=100$, $T=4$ and $\mathbf{x}_{it}=x_i\overset{iid}{\sim} N(0,1)$ for all $i$ and $t$. For the association structure, assume that the random variables $\mathbf{e}_i^{B}=(e^{B}_{i1},e^{B}_{i2},e^{B}_{i3},e^{B}_{i4})^{\prime}$ in \@ref(eq:TMB) are iid random vectors from the tetra-variate normal distribution with mean vector the zero vector and covariance matrix the correlation matrix $\mathbf{R}$ given by
\begin{equation}
\mathbf{R}=\left( {\begin{array}{*{20}c}
   1.00 & 0.90 & 0.90 & 0.90  \\
   0.90 & 1.00 & 0.90 & 0.90  \\
   0.90 & 0.90 & 1.00 & 0.90 \\
   0.90 & 0.90 & 0.90 & 1.00
 \end{array} } \right).
 (\#eq:CMB)
 \end{equation}
This association configuration defines an exchangeable correlation matrix for the clustered binary responses, i.e. $\text{corr}(Y_{it_1},Y_{it_2})=\rho_i$ for all $i$ and $t$. The strength of the correlation ($\rho_i$) is decreasing as the absolute value of the time-stationary covariate $x_i$ increases. For example, $\rho_i=0.7128$ when $x_{i}=0$ and $\rho_i=0.7$ when $x_i=3$ or $x_i=-3$. Therefore, a strong exchangeable correlation pattern for each subject that does not differ much across subjects is implied with this configuration.
```

```{r}
set.seed(123)
# sample size
N <- 100
# cluster size
clsize <- 4
# intercept
intercepts <- 0
# regression parameter associated with the covariate
betas <- 0.2
# correlation matrix for the NORTA method
cor.matrix <- toeplitz(c(1, 0.9, 0.9, 0.9))
# time-stationary covariate
x <- rep(rnorm(N), each = clsize)
# simulation of clustered binary responses
CorBinRes <- rbin(clsize = clsize, intercepts = intercepts, betas = betas, 
    xformula = ~x, cor.matrix = cor.matrix, link = "probit")
library(gee)
# fitting a GEE model
binGEEmod <- gee(y ~ x, family = binomial("probit"), id = id, data = CorBinRes$simdata)
# checking the estimated coefficients
summary(binGEEmod)$coefficients
```
\textbf{ }
```{example, label=BinaryLogit, name="Simulation of clustered binary responses under a conditional marginal logit model without utilizing the NORTA method"} 
Consider now simulation of correlated binary responses from the marginal logit model 
\begin{equation*}
\Pr(Y_{it}=1 |\mathbf{x}_{it})=F(0.2x_i)
\end{equation*}
where $F$ is the cdf of the standard logistic distribution (mean $=0$ and variance $=\pi^2/3$), $N=100$, $T=4$ and $\mathbf{x}_{it}=x_i\overset{iid}{\sim} N(0,1)$ for all $i$ and $t$. This is similar to the marginal model configuration in Example \@ref(exm:BinaryProbit) except from the link function. For the dependence structure, assume that the correlation matrix of $\mathbf{e}_i^{B}=(e^{B}_{i1},e^{B}_{i2},e^{B}_{i3},e^{B}_{i4})^{\prime}$ in \@ref(eq:TMB) is equal to the correlation matrix $\mathbf{R}$ defined in \@ref(eq:CMB). To simulate $\mathbf{e}_i^{B}$ without utilizing the NORTA method, one can employ the tetra-variate extreme value distribution [@Gumbel1958]. In particular, this is accomplished by setting $\mathbf{e}_i^{B}=\mathbf{U}_i-\mathbf{V}_i$ for all $i$, where $\mathbf{U}_i$ and $\mathbf{V}_i$ are independent random vectors from the tetra-variate extreme value distribution with dependence parameter equal to $0.9$, that is
$$\Pr\left(U_{i1}\leq u_{i1},U_{i2}\leq u_{i2},U_{i3}\leq u_{i3},U_{i4}\leq u_{i4}\right)=\exp\left\{-\left[\sum_{t=1}^4 \exp{\left(-\frac{u_{it}}{0.9}\right)}\right]^{0.9}\right\}$$
and 
$$\Pr\left(V_{i1}\leq v_{i1},V_{i2}\leq v_{i2},V_{i3}\leq v_{i3},V_{i4}\leq v_{i4}\right)=\exp\left\{-\left[\sum_{t=1}^4 \exp{\left(-\frac{v_{it}}{0.9}\right)}\right]^{0.9}\right\}.$$
It follows that $e_{it}^{B}\sim F$ for all $i$ and $t$ and $\textrm{corr}(\mathbf{e}_i^{B})=\mathbf{R}$ for all $i$. 
```

```{r}
set.seed(8)
# simulation of epsilon variables
library(evd)
rlatent1 <- rmvevd(N, dep = sqrt(1 - 0.9), model = "log", d = clsize)
rlatent2 <- rmvevd(N, dep = sqrt(1 - 0.9), model = "log", d = clsize)
rlatent <- rlatent1 - rlatent2
# simulation of clustered binary responses
CorBinRes <- rbin(clsize = clsize, intercepts = intercepts, betas = betas, 
    xformula = ~x, rlatent = rlatent)
# fitting a GEE model
binGEEmod <- gee(y ~ x, family = binomial("logit"), id = id, data = CorBinRes$simdata)
# checking the estimated coefficients
summary(binGEEmod)$coefficients
```

## No marginal model specification
To achieve simulation of clustered binary, ordinal and nominal responses under no marginal model specification, perform the following intercepts:

1. Based on the marginal probabilities calculate the intercept of a marginal probit model for binary responses (see Example \@ref(exm:no-covariate)) or the category-specific intercepts of a cumulative probit model [see third example in @Touloumis2016] or of a baseline-category logit model for multinomial responses (see Example \@ref(exm:Example2NoCovariates)).

2. Create a pseudo-covariate say `x` of length equal to the number of cluster size (`clsize`) times the desired number of clusters of simulated responses (say `R`), that is `x = clsize * R`. This step is required in order to identify the desired number of clustered responses.  

3. Set `betas = 0` in the core functions `rbin` (see Example \@ref(exm:no-covariate)) or `rmult.clm`, or set 0 all values of the `beta` argument that correspond to the category-specific parameters in the core function `rmult.bcl` (see Example \@ref(exm:Example2NoCovariates)).

4. set `xformula = ~ x`.

5. Run the core function to obtain realizations of the simulated clustered responses.

\textbf{ }

```{example, label=no-covariate, name="Simulation of clustered binary responses without covariates"}
Suppose the goal is to simulate $5000$ clustered binary responses with $\Pr(Y_{t}=1)=0.8$ for all $t=1,\ldots,4$. For simplicity, assume that the clustered binary responses are independent.
```

```{r}
set.seed(123)
# sample size
N <- 5000
# cluster size
clsize <- 4
# intercept
intercepts <- qnorm(0.8)
# pseudo-covariate
x <- rep(0, each = clsize * N)
# regression parameter associated with the covariate
betas <- 0
# correlation matrix for the NORTA method
cor.matrix <- diag(clsize)
# simulation of clustered binary responses
CorBinRes <- rbin(clsize = clsize, intercepts = intercepts, betas = betas, 
    xformula = ~x, cor.matrix = cor.matrix, link = "probit")
library(gee)
# simulated marginal probabilities
colMeans(CorBinRes$Ysim)
```

\textbf{ }
```{example, Example2NoCovariates, name="Simulation of clustered nominal responses without covariates"}
Suppose the aim is to simulate $N=5000$ clustered nominal responses with
$\Pr(Y_{t}=1)=0.1$, $\Pr(Y_{t}=2)=0.2$, $\Pr(Y_{t}=3)=0.3$ and $\Pr(Y_{t}=4)=0.4$, for all $i$ and $t=1,\ldots,3$. For the sake of simplicity, we assume that the clustered responses are independent. 
```

```{r, tidy=TRUE}
# sample size
N <- 5000
# cluster size
clsize <- 3 
# pseudo-covariate 
x <- rep(0, each = clsize * N)
# parameter vector
betas <-  c(log(0.1/0.4), 0, log(0.2/0.4), 0, log(0.3/0.4), 0, 0, 0)
# number of nominal response categories
ncategories <- 4  
set.seed(1) 
# correlation matrix for the NORTA method
cor.matrix <- kronecker(toeplitz(c(1,rep(0,clsize-1))),diag(ncategories)) 
# simulation of clustered nominal responses
CorNorRes <- rmult.bcl(clsize = clsize, ncategories = ncategories, betas = betas, xformula = ~x, cor.matrix = cor.matrix)
# simulated marginal probabilities
apply(CorNorRes$Ysim,2,table)/N
```
# How to Cite
```{r, comment=""}
citation("SimCorMultRes")
```

# References
